---
title: "Lung Function Analysis Stat 741 Group Project"
author: "Yifei Xu, Shubh Patel, Max Brown"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
---

## Introduction

In this report we... Fix this
Research Question
1.  Do asthmatics have steeper rates of decline (slope) or lower levels of lung function (FEV1) than non-asthmatics, independent of smoking history? 

2. Do asthmatic smokers have steeper rates of decline or lower levels of lung function than non-asthmatic smokers? Here you can explore three options 1) using the fixed effect variable ever smoke, 2) using the time dependent variable for smoking, or 3) using the current number of cigarettes per day. When using this I would suggest making it a categorical variable using quartiles etc.

## Data
Data stuff will go here
```{r setup, include=FALSE}
library(tidyverse)
library(haven)
```

```{r}
# Load the data
asthma_df <- read_dta("data/family asthma p1 data version2.dta")
```

```{r}
# Reshape the data
asthma_tidy_df <- asthma_df %>%
  pivot_longer(
    cols = matches("\\d+$"),           
    names_to = c(".value", "visit"),  
    names_pattern = "([a-z]+)(\\d+)"   
  ) %>%
  select(family, id, visit, everything())

```


## Data Cleaning
Data cleaning will go here

## Exploratory Analysis
Analysis here

## Models

### Model 1

### Model 2

## Model Comparisons

## Diagnostics

## Conclusions





